Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-03 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Remove or rewrite parts in bold face until results

Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous, but treating it can also be tedious. If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to *impute* (i.e., fill in) the missing values in the incomplete observations. Subsequently, statistical inferences can be performed on the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as 'multiple imputation' [MI; @rubin76]. MI has proven to be a powerful technique to yield unbiased and confidence valid estimates of the true---but missing---data inference under many circumstances [@buur18].

Figure \ref{fig:diagram} provides an overview of the steps involved with MI---from incomplete data, to $M$ multiply imputed datasets, to $M$ estimated quantities of interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$. Missing data in dataset $y$ is imputed $M$ times. The imputed data $y_{imp}$ is combined with the observed data $y_{obs}$ to create $M$ completed datasets. On each completed dataset, the analysis of scientific interest is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $M$ separate $\hat{Q}$-values are obtained. These $M$ values are combined into a single pooled estimate $\bar{Q}$.

@@

```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation."}
knitr::include_graphics("./images/diagram.pdf") #adapted from [@buur18] par. 1.4.1.
```

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values. This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data? These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. Moreover, there is no scientific consensus on how to evaluate convergence of MI algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.

Currently, the recommended practice for evaluating convergence in iterative imputation algorithms is to visually inspect imputations for signs of non-convergence. **Add that this refers to chain means and variances already? It's insufficient that these are univariate, so multivariate state space of the algorithm may not be converged when univariates are ok. And @buur18's suggestion to track user-defined statistics (e.g., regression coefficient) may be somewhat advanced for empirical researchers. Moreover, this is not model-independent, i.e., only apply to one substantive model, while one of the advantages of MI is that missing data problem and scientific problem are split. Then we can say 'we propose a novel...' and integrate this: Note that usually we only evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). With these we cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). Van Buuren [-@buur18, $\S$ 4.5.2] proposed to implement multivariable evaluation of the MICE algorithm through eigenvalue decomposition building on the work of @mack03. Eigenvalues of a covariance matrix are a measure of the data's covariance.**.

This method is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2]. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---yet not straightforward. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics we may only identify signs of non-convergence [@hoff09]. Several of such diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed. For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R]. The convergence properties of the MICE algorithm are investigated through model-based simulation. The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets. **add societal relevance: what is the consequence of non-convergence?**

**Include sub-questions and add structure of the paper!** How can non-convergence be identified diagnostically? Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? **How severe is it when the algorithm has not converged? And what are guidelines for practice? Can the parameter of interest, estimand $Q$, be correct when the algorithm is not (yet) converged, and vice versa?** *(Maybe add these too? What are the effects of continued iterations on estimates, predictions and inferences? Do the answers differ with varying missingness proportions? That is, we vary the nr of iterations and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)*

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. **The number of units $i$ in dataset $y$ with at least one missing data value, divided by the total number of units $n$, is called the missingness proportion in $p_{mis}$.** The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$). The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. **Add: $\theta$s are scalar summaries of interest in the iterative algorithm (e.g., chain means; the average of the imputed values in each imputation chain). **

@@

**The goal of imputation through e.g. MICE is to obtain a multivariate distribution by iterating over the sequence of univariate imputations. Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations, or by tracking the $\widehat{R}$ and autocorrelation. The downside to these approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Focusing on the convergence of these outcome parameters may influence this procedure in the sense that the model of evaluation favors the model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. We propose eigenvalue as such a parameter, because it.... **

Mixing and stationarity have histroically been inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances). As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient). A user-specified scalar summary, however, is not universal to all complete data problems. Therefore, as inspired by [@mack03], we propose a $\theta$ that summarizes the multivariate state space of the algorithm. Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets. **The first eigenvalue has the appealing property that is not dependent on the substantive model of interest.**

**Since convergence of iterative imputation algorithms is in distribution, there is not a unique point at which the algorithm reaches a converged state. We can only evaluate scalar summaries of the multivariate state space of the algorithm, $\theta$s. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2]. Because of the issues mentioned above, we propose a novel $\theta$ to monitor: the first eigenvalue of the variance-covariance matrix of the completed data. Let $\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$. $\lambda_1$ is measure that summarises the covariances in the completed datasets.**

There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13]. Without mixing, imputation chains do not intermingle nicely, **indicating that ...**. Without stationarity, there is trending within imputation chains, which implies that further iterations would yiled a different set of imputations.

To illustrate what non-convergence looks like in MI, we reproduce the example from van Buuren [-@buur18, $\S$ 6.5.2]. Figure \ref{fig:non-conv} shows the traceplot for one of the variables in the example. we see the average of the imputed values for a variable $j$ in $y_{imp}$. The average imputed values in the left hand plot are about a magnitude 2 lower than the right hand plot. So non-convergence has a impact on the average imputed value for $j$ in $y_{imp, m}$. This difference (presumably) translates to bias in the pooled estimate $\bar{\mu}_j$ as well. And maybe also to higher order statistics, e.g. variances, covariances, etc. Therefore, it's important to converge. **Explain what we see, namely example by @buur18 reproduced, showing the traceplots of chain means for some variable. The first plot is typical convergence of MICE, the second is pathological non-convergence because of a mis-specified imputation model. Each line is an imputation. In the first plot, the chains intermingle nicely and there is little to no trending. In the second plot, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a magnitude two larger than those of the typically converged algorithm. This shows the importance of reaching converged states in iterative imputation algorithms.**

## Methods under evaluation

@@

Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92]. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, use we $\widehat{R}$ and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96.
<!-- @cowl96, p. 898. -->

```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence. Please note that this is the same data with a different imputation model, leading to different imputations (see the y axis!). Wel min of meer dezelfde origin op iteration 1."}
load("./images/nonc_plot.Rdata")
load("./images/conv_plot.Rdata")
par(mar = c(2, 2, 0.1, 0.1))
plot(conv)
plot(nonc)

@@

An updated version of $\widehat{R}$ has been proposed by @veht19 (p. 5). **This version may be suitable for iterative imputation.** Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance). For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains. Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one another, which may increase the variance of $\theta$ and speed up convergence. As convergence diagnostic, the interest is therefore in positive $AC$-values. **Maybe remove:** Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of our scope.

@@

Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution. In practice, non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01. And a t-test is performed to assess whether $AC$ is significantly different from zero.

We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed. Because even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity. Upon *sufficient* convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there may be some dependency between subsequent iterations of imputation chains ($AC > 0$).

**Before using them in the simulation, the two methods show at least be able to distinguish between the two algorithms plotted in Figure 2. From visual inspection we know that the typical convergence has some signs of non-mixing around iteration 2, and little trending. In the non-convergence situation, there is a lot of trending upto iteration 6, after which the chains reach a somewhat more stationary state. Mixing in this situation gets worse from the first iteration onward, and gradually gets better around iteration 7. To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence identifiers for iterative imputation algorithms, the following condition must hold: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance). And additionally the methods should reflect the increasing convergence in the typical convergence situation as the number of iterations goes up.**

**In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again, now together---as are the non-convergence diagnostics. Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations. As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up **in the typical convergence situation**. The superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$. Panel C displays the $AC$ as computed with the R function `stats::acf()`. When we look at this panel, we conclude something weird. The $AC$-values indicate equal performance (up-to $t=5$) for the typical convergence and the pathological non-convergence, while there is obvious trending in the latter. Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven. After consulting the documentation on stats::acf() [@R], we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms. There is a correction factor for a mathematical shortcut that works in the limit. According to @box15, the function works when the number of iterations > 50. The default number of iterations in iterative imputation, however, is often much lower. Therefore, we compute $AC$ manually, see panel D. The $AC$ values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**

**acf is niet bedoeld om weinig iteraties te... niet een fout in acf, alleen hier niet van toepassing**

@@

@@

The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest $Q$. Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence (**add thetas**). And finally, we will formulate an informed advice on the requirements to safely assume *sufficient* convergence in practice. That is, we assume that the algorithm is *sufficiently* converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$. (**add to formulate advice on thetas?**)

@@

- Estimate $\hat{Q}$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. Then pool the $\hat{Q}$s with `base::mean()` to get $\bar{Q}$s.

- Compute $\lambda_1$, to use as $\theta$ in convergence diagnostics. By definition, the first eigenvalue of the variance-covariance matrix, $\lambda_1$, is equal to the variance of the principal component solution of each {$y_{obs}, y_{imp, m}$}, which we calculate with `stats::princomp()` [@R].

>>>>

Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous, but treating it can also be tedious. If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to *impute* (i.e., fill in) the missing values in the incomplete observations. Subsequently, statistical inferences can be performed on the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as 'multiple imputation' [MI; @rubin76]. MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances [@buur18].

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. Missing data $y_{mis}$ in an incomplete dataset is imputed $M$ times. The imputed data $y_{imp}$ is combined with the observed data $y_{obs}$ to create $M$ completed datasets. On each completed dataset, an analysis of scientific interest is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with $Q$. Since $Q$ is estimated on each completed dataset, $M$ separate $\hat{Q}$-values are obtained. These $M$ values are combined into a single pooled estimate $\bar{Q}$. The premise of multiple imputation is that $\bar{Q}$ is an unbiased and confidence valid estimate of the true---but missing---data inference.

@@

```{r diagram, out.width="\\linewidth", include=TRUE, fig.cap="Scheme of the main steps in multiple imputation (where $M=3$)---from an incomplete dataset, to $M$ multiply imputed datasets, to $M$ estimated quantities of scientific interest $\\hat{Q}$s, to a single pooled estimate $\\bar{Q}$."}
knitr::include_graphics("./images/diagram.pdf") #adapted from [@buur18] par. 1.4.1.
```

A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. With MICE, imputed values $y_{imp}$ are drawn from the posterior predictive distribution of the missing values $y_{mis}$. The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data? These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. Moreover, there is no scientific consensus on how to evaluate convergence of imputation algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.

Currently, the recommended practice for evaulating the convergence of iterative imputation algorithms is through visual inspection. That is, the state space of the algorithm is monitored by plotting a scalar summary $\theta$ against the iteration number.
**[Segway to WHAT TO MONITOR? Also, this paragraph is more or less the same as the third paragraph in the second section. Merge or remove one of the two?]** Typically, the $\theta$s under evaluation are chain means and variances (**define!**). These $\theta$s may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of $y_{imp}$. A suggestion by @buur18 for a multivariate $\theta$ to monitor is the quantity of scientific interest $Q$ (e.g., a regression coefficient). Implementing this, however, might be somewhat advanced for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. Therefore, this $\theta$ is not sufficient either. Van Buuren [-@buur18, $\S$ 4.5.2] also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of @mack03. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. This would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented. 
<!-- Recommendation: "the different streams should be freely intermingled with one another, without showing any definite trends". -->
**[Segway to HOW TO MONITOR?]** Monitoring convergence visually is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2]. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics  may only identify signs of *non*-convergence [@hoff09]. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed. First, we define which non-convergence diagnostics will be considered, and evaluate how these diagnostics could be appropriate for iterative imputation applications. In the second part of this paper, we will investigate the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in `R` [@R]. The aim of the simulation is to determine whether unbiased, confidence valid inferences may be obtained under different levels of non-convergence (**and at which point convergence may safely be assumed**). Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence. The results of this study are guidelines for identifying non-convergence in iterative imptation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets. **[add more societal relevance: what is the consequence of non-convergence? e.g., biased estimates, invalid inference, etc.]** For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular `mice` package [@mice] in `R` [@R].

**[Include sub-questions and make structure of the paper clear!]** How can non-convergence be identified diagnostically? Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? How severe is it when the algorithm has not converged? And what are guidelines for practice? Can the parameter of interest $Q$ be correct when the algorithm is not (yet) converged, and vice versa?

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ in dataset $y$ with at least one missing value, divided by the total number of units $n$, is called the missingness proportion $p_{mis}$. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ plausible values, where $M$ is the number of imputations ($m = 1, 2, \dots, M$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm. The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. **$\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain). **

@@

There are two requirements for convergence of iterative algorithms: mixing and stationarity [@gelm13]. Without mixing, imputation chains do not intermingle nicely, indicating that the distribution of imputed values differs across imputations. Chains may be 'stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values. This may cause under-estimation of the variance within chains, which results in spurious, invalid inferences. Without stationarity, there is trending within imputation chains. Trending implies that further iterations would yield a different set of imputations. Iterative imputation algorithms that have not yet reached stationarity, may thus yield biased estimates.

<!-- **The goal of imputation through e.g. MICE is to obtain a multivariate distribution by iterating over a sequence of univariate imputations. Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations. The downside to these approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Focusing on the convergence of these outcome parameters may influence this procedure in the sense that the model of evaluation favors the model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. We propose eigenvalue as such a parameter, because it.... ** -->

**[Move to INTRO???]** To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2]. Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call $j$. We see two traceplots of the chain means for $j$---i.e., the average of the imputed values for $j$ in each imputation $y_{imp,m}$, plotted against the iteration number. Each line portrays one imputation. The plot on the lefthand side of the figure shows typical convergence of an iterative imputation algorithm. The righthand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. **[Leave this in if it's after mixing and stationarity are defined, otherwise remove?] In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence plot, there is a lot of trending and some chains do not intermingle.** Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. That means that non-convergence has a impact on the mean of $j$ in $y_{imp, m}$. This difference (presumably) translates to bias in the $M$ completed data estimates $\hat{\mu}_{j,m}$, and consequently also to the pooled estimate $\bar{\mu}_j$. Non-convergence thus leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

```{r non-conv, out.width='.49\\linewidth', fig.width=3.5, fig.height=3,fig.show='hold',fig.cap="Typical convergence versus pathological non-convergence. Please note that this is the same data with a different imputation model, leading to different imputations (e.g., see the units on the y-axis)."}
load("./images/nonc_plot.Rdata")
load("./images/conv_plot.Rdata")
par(mar = c(2, 2, 0.1, 0.1))
plot(conv)
plot(nonc)
```

**[Move to INTRO???]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteraton number is plotted against a certain $\theta$. $\theta$s are scalar summaries of the state space of the algorithm at a specific iteration. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2]. As @buur18 describes, researchers can specify their quantity of scientific interest $Q$ (e.g., a regression coefficient) as scalar summary $\theta$. Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest.  The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a $\theta$ that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose $\lambda_1$ as such a scalar summary. We define $\lambda_1$ as the first eigenvalue of the variance-covariance matrix of the completed data. Let $\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$. $\lambda_1$ is measure that summarises the covariances in the completed datasets. **The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by @mack03.**

<!-- **Since convergence of iterative imputation algorithms is in distribution, there is not a unique point at which the algorithm reaches a converged state. We can only evaluate scalar summaries of the multivariate state space of the algorithm, $\theta$s. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2]. Because of the issues mentioned above, we propose a novel $\theta$ to monitor: the first eigenvalue of the variance-covariance matrix of the completed data. Let $\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $M$ completed datasets ${y_{obs}, y_{imp}}$. $\lambda_1$ is measure that summarises the covariances in the completed datasets.**  -->

## Diagnostics under evaluation

@@

Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for converegnce: mixing or stationarity. Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92]. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, we use $\widehat{R}$ and $AC$ to evaluate mixing and stationarity separately, as recommended by e.g., @cowl96.
<!-- @cowl96, p. 898. -->

@@

An updated version of $\widehat{R}$ has been proposed by Vehtari et al. [-@veht19, p. 5]. **This version may be suitable for iterative imputation.** Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain means). For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains. Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one another, which may increase the variance of $\theta$ and speed up convergence. As convergence diagnostic, the interest is therefore in positive $AC$-values. The magnitude of $AC$-values may be evaluated statistically [**add reference**].

### Thresholds

<!-- We assume that complete convergence, defined as $\widehat{R} = 1$ and $AC = 0$, will not be observed.  -->
Upon convergence, $\widehat{R}=1$ and $AC=0$, which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution. Even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity. Upon *sufficient* convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} > 1$), and there may be some dependency between subsequent iterations of imputation chains ($AC > 0$). In practice, non-convergence is usually diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01. And a t-test is performed to assess whether $AC$ is significantly different from zero.

@@

Before we evaluate the performance of the non-convergence diagnostics $\widehat{R}=1$ and $AC=0$ quantitatively through simulation, we apply them to the example of pathological non-convergence by @buur18 that we reproduced above. We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms. But $\widehat{R}=1$ and $AC=0$ should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}. From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around $t=2$), but intermigled nicely overall. Additionally, there was very little trending. The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from $t=7$). Moreover, there was a lot of trending as well initially (up-to $t=6$), after which the chains reached a somewhat more stationary state. To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence identifiers for iterative imputation algorithms, they should identify the same trends as the visual inspection. But at least, the following condition must hold: the methods should indicate worse performance for the mis-specified model with pathological non-convergence than the typically converged algorithm (i.e., higher $\widehat{R}$- and $AC$-values).

In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again---now together. Panel B shows two versions of calculating $AC$s applied to the $\theta$s of panel A: the default calculation with `R` function `stats::acf()` [@R], and manual calculation as the correlation between $\theta$ in iteration $t$ and $\theta$ in iteration $t+1$. Panel C shows the traditional computation of $\widehat{R}$, and panel D shows $\widehat{R}$ as computed by implementing @veht19 's recommendations.

@@

When we look at panel B, we conclude something weird. The $AC$-values calculated with the default function indicate equal performance for the typical convergence and the pathological non-convergence (up-to $t=5$), while there is obvious trending in the $\theta$s of the latter. Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven. After consulting the documentation on `stats::acf()` [@R], we conclude that this $AC$ function is not suited for iterative imputation algorithms. The function is optimized for performance when $t\geq50$ [@box15], while the default number of iterations in iterative imputation is often much lower. Therefore, we compute $AC$ manually, see the solid line. These $AC$ values indicate non-stationarity as expected. We therefore calculate $AC$ manually.

**Describe panel C here and why it is not optimal]** The $\widehat{R}$-values in panel D do meet the requirements specified above. $\widehat{R}$ as computed conform @veht19 indeed indicate less signs of non-convergence as the number of iterations goes up. 
<!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->

@@

The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest $Q$. Subsequently we will evaluate how well $\widehat{R}$ and $AC$ perform in identifying the effects of non-convergence **[and performance of different thetas]**. And finally, we will formulate an informed advice on the requirements to safely assume *sufficient* convergence in practice. That is, we assume that the algorithm is *sufficiently* converged when each estimate $\bar{Q}$ is an unbiased and confidence valid estimate of the corresponding estimand $Q$. **[and formulate advice on thetas?]**

@@

- Estimate $\hat{Q}$s for $Q= \mu$ and $Q=\sigma$ with `base::mean()`, and `stats::sd()` [@R]. We pool the $\hat{Q}$s with `base::mean()` to get $\bar{Q}$s.

- Compute $\lambda_1$, to use as $\theta$ in the two convergence diagnostics. By definition, the first eigenvalue of the variance-covariance matrix, $\lambda_1$, is equal to the variance of the principal component solution of each {$y_{obs}, y_{imp, m}$}, which we calculate with `stats::princomp()` [@R].
